---
title: "Getting Started with np"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with np}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(np.messages = FALSE)
```

This vignette is meant to be the smallest useful package-side introduction to
`np`. The emphasis is on one clean workflow that users can run after
installation: choose a bandwidth, fit a model, inspect the result, and plot it.

Broader worked examples, package comparisons, and method-specific articles are
better carried by the gallery site:

- <https://jeffreyracine.github.io/gallery/np_npRmpi.html>
- <https://jeffreyracine.github.io/gallery/quickstarts.html>

## The basic workflow

In `np`, the bandwidth object is often the key object in the analysis.

1. compute or inspect a bandwidth object,
2. fit the model,
3. summarize or plot the result.

## A simple regression example

```{r}
library(np)
data(cps71, package = "np")

bw <- npregbw(logwage ~ age, data = cps71)
summary(bw)

fit <- npreg(bws = bw)
summary(fit)
```

## Plotting the fitted relationship

```{r, fig.width = 6, fig.height = 4}
plot(cps71$age, cps71$logwage, cex = 0.25, col = "grey")
lines(cps71$age, fitted(fit), col = 2, lwd = 2)
```

## Predict and request uncertainty without repeating search

Use the retained bandwidth object when only the evaluation points or requested
outputs change. Keep the original training data available to the fitted call.

```{r}
new_points <- data.frame(age = c(25, 40, 55))
predict(fit, newdata = new_points)

# This refits at the existing bandwidth; it does not run bandwidth selection.
fit_inference <- npreg(bws = fit$bws, gradients = TRUE, se = TRUE)
head(se(fit_inference))
head(gradients(fit_inference))
head(gradients(fit_inference, se = TRUE))

prediction <- predict(fit, newdata = new_points, se.fit = TRUE)
prediction$fit
prediction$se.fit
```

The switches and extractors have different roles:

| Request | Purpose |
| --- | --- |
| `se = TRUE` in an estimator call | Compute its supported fit-time uncertainty. |
| `gradients = TRUE` in an estimator call | Compute supported derivatives or categorical contrasts. Add `se = TRUE` for their standard errors. |
| `se(fit)` or `gradients(fit, se = TRUE)` | Read stored standard errors; do not refit or repeat search. |
| `predict(fit, newdata = ..., se.fit = TRUE)` | Evaluate predictions and their supported standard errors. |
| `vcov(fit)` | Read coefficient covariance for families such as `npindex` and `npplreg`, not a universal covariance of the fitted surface. |
| `plot(fit, errors = "bootstrap")` or `errors = "asymptotic"` | Request plot-time uncertainty through that family's plot method. |

Consult the estimator's help for supported kernels, bandwidth types and
derivative orders. Uncertainty not requested at fitting produces an extractor
error with a no-search refit hint. This differs from an unsupported inference
configuration or an undefined evaluation row (for example, no compact-kernel
support). Inspect warnings and `NA` results rather than treating them as zero
uncertainty. Asymptotic results condition on the selected smoothing settings;
they do not include bandwidth or polynomial-degree selection uncertainty.
Plot bands apply within a panel, not jointly across all displayed panels;
asymptotic simultaneous bands are not available. See `?np.plot` for band and
bootstrap choices.

## Moving from 0.70-5 to 0.80-1

Fit-time standard errors are now opt-in (`se = TRUE`) across the principal
families. The intentional exception is `npindex`, whose default
`se = TRUE, se.type = "asymptotic"` includes coefficient covariance;
`se = FALSE` requests a point-only fit. Gradients remain a separate request.

Use current dotted argument names such as `gradient.order`, `data.overlay`
and `data.rug`, and `B` for bootstrap replication counts where documented.
Unknown named arguments are rejected with a spelling hint where available;
an old misspelling that was previously ignored can therefore surface now.
The full package NEWS (`news(package = "np")`) lists the replacements and
family-specific changes. The [Gallery FAQ](https://jeffreyracine.github.io/gallery/faq.html)
provides further troubleshooting guidance. Refit using `fit$bws` as above when the missing
piece is uncertainty or gradients, not a new bandwidth choice.

## Mixed data

One important feature of `np` is that it handles mixed data directly. Variable
class matters: unordered categorical variables should be factors, and ordered
categorical variables should be ordered factors when appropriate.

```{r}
set.seed(42)
mydat <- data.frame(
  y = rnorm(200),
  x_cont = runif(200),
  x_unordered = factor(sample(c("a", "b", "c"), 200, replace = TRUE)),
  x_ordered = ordered(sample(1:4, 200, replace = TRUE))
)

bw_mixed <- npregbw(y ~ x_cont + x_unordered + x_ordered, data = mydat)
fit_mixed <- npreg(bws = bw_mixed)
summary(fit_mixed)
```

## A note on modern local-polynomial search

For local-polynomial-capable methods, `np` now supports joint selection of
polynomial order and bandwidth. The modern route is to use
`search.engine = "nomad+powell"` when you want the search to choose both
together.

If you want the recommended route without spelling out all of the LP tuning
arguments, use `nomad = TRUE`. This is a documented convenience preset, not a
generic optimizer alias: it fills only missing values among the LP degree-search
controls and leaves compatible explicit overrides in place. This route uses the
optional NOMAD backend provided by the suggested package `crs`, so install
`crs` first if you want to use `nomad = TRUE` or
`search.engine = "nomad"`/`"nomad+powell"`.

```{r}
if (requireNamespace("crs", quietly = TRUE) &&
    utils::packageVersion("crs") >= package_version("0.15-41")) {
  set.seed(7)
  n <- 120
  x <- runif(n, -1, 1)
  y <- x + 0.4 * x^2 + rnorm(n, sd = 0.18)

  fit_nomad <- npreg(y ~ x, nomad = TRUE, degree.max = 1L, nmulti = 1L)
  fit_nomad$bws$nomad.shortcut

  # Tune one component explicitly while leaving the rest of the preset in place.
  fit_nomad_direct <- npreg(
    y ~ x,
    nomad = TRUE,
    search.engine = "nomad",
    degree.max = 1L,
    nmulti = 1L
  )
}
```

The same convenience entry point is available for the other LP-capable families:
`npcdens`, `npcdist`, `npplreg`, `npscoef`, and `npindex`, together with their
corresponding `*bw` constructors.

Keep the first run modest and runnable. Fuller worked examples belong on the
gallery rather than in this package vignette.

## Data preparation matters

In `np`, the formula interface tells the function which variables are the
response and regressors. It is not imposing an ordinary linear-additive model.

It is also important not to pass blocks of 0/1 dummies as if this were a
standard linear-model workflow. If the underlying variable is categorical, it
is usually better to keep it as one `factor` or `ordered` variable.

## Other common starting points

This vignette keeps the package-side introduction intentionally narrow. Other
common first routes are:

- `?npudens` and `?npudist` for unconditional density and distribution work,
- `?npcdens`, `?npcdist`, and `?npqreg` for conditional density, distribution,
  and quantiles,
- `?npconmode` for classification and conditional mode estimation,
- `?npplreg`, `?npindex`, and `?npscoef` for semiparametric models.

Those broader branches are better carried by help pages and website articles
than by a single shipped vignette.

## Where to go next

- `vignette("np_entropy_tests", package = "np")` for a compact package-side
  testing overview
- `?npreg`, `?npregbw`, `?npudens`, and `?npcdens` for core help pages
- <https://jeffreyracine.github.io/gallery/kernel_primer.html> for the
  conceptual kernel overview
- <https://jeffreyracine.github.io/gallery/density_distribution_quantiles.html>
  for density, distribution, and quantile workflows
- <https://jeffreyracine.github.io/gallery/semiparametric_models.html> for
  partially linear, single-index, and varying-coefficient routes
