---
title: "Getting Started with npRmpi"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with npRmpi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  purl = FALSE
)
```

This vignette is intentionally conservative. The goal is to provide a
package-side overview of the modern `npRmpi` workflow without making routine
vignette builds depend on live MPI startup.

The full install matrix, platform-specific notes, and longer mode comparisons
belong on the gallery site rather than in a shipped package vignette:

- <https://jeffreyracine.github.io/gallery/mpi_large_data.html>
- <https://jeffreyracine.github.io/gallery/quickstarts.html>

The main practical point is simple: on platforms where spawning is supported,
the recommended interactive route is now `session` mode (the `spawn` code
path), where the estimator calls look like ordinary `np` code once MPI has been
initialized with `npRmpi.init(nslaves = ...)`.

## Which mode should I use?

- Use `session` / `spawn` for ordinary interactive work on platforms where
  spawning is supported.
- Use `attach` when the MPI world is already launched with `mpiexec` and you
  still want `npRmpi.init(mode = "attach")` to manage the worker loop.
- Use `profile` / manual-broadcast mode when startup and broadcast logic need to
  be explicit, especially on larger or heterogeneous clusters.

For teaching and troubleshooting, it is usually better to be explicit than to
rely on `mode = "auto"`.

## Recommended starting point: session mode

```{r, purl = FALSE}
library(npRmpi)

npRmpi.init(nslaves = 1)

set.seed(1)
x <- runif(200)
y <- sin(2 * pi * x) + rnorm(200, sd = 0.2)

bw <- npregbw(y ~ x, regtype = "ll", bwmethod = "cv.ls")
fit <- npreg(bws = bw)
summary(fit)

new_points <- data.frame(x = c(0.25, 0.5, 0.75))
predict(fit, newdata = new_points)

# Reuse the bandwidth: no second bandwidth search is performed.
fit_inference <- npreg(bws = fit$bws, gradients = TRUE, se = TRUE)
head(se(fit_inference))
head(gradients(fit_inference))
head(gradients(fit_inference, se = TRUE))
prediction <- predict(fit, newdata = new_points, se.fit = TRUE)
prediction$fit
prediction$se.fit

npRmpi.quit()
```

The examples in this vignette are alternative session layouts, not one script
to concatenate after MPI has finalized. Keep the original training data
available when refitting or predicting with a retained bandwidth object.

## Uncertainty and migration from 0.70-5

Statistical requests follow the same conventions as `np`:

| Request | Purpose |
| --- | --- |
| `se = TRUE` in an estimator call | Compute its supported fit-time uncertainty. |
| `gradients = TRUE` in an estimator call | Compute supported derivatives or categorical contrasts. Add `se = TRUE` for their standard errors. |
| `se(fit)` or `gradients(fit, se = TRUE)` | Read stored standard errors; do not refit or repeat search. |
| `predict(fit, newdata = ..., se.fit = TRUE)` | Evaluate predictions and their supported standard errors. |
| `vcov(fit)` | Read coefficient covariance for families such as `npindex` and `npplreg`, not a universal covariance of the fitted surface. |
| `plot(fit, errors = "bootstrap")` or `errors = "asymptotic"` | Request plot-time uncertainty through that family's plot method. |

Fit-time standard errors are opt-in across the principal families. The
intentional exception is `npindex`, whose default
`se = TRUE, se.type = "asymptotic"` includes coefficient covariance;
`se = FALSE` requests a point-only fit. Gradients remain a separate request.
The example above shows the no-search refit used when stored results are absent.

Consult each estimator's help for supported kernels, bandwidth types and
derivative orders. Missing stored uncertainty, unsupported inference and an
undefined evaluation row are different conditions; do not interpret `NA` as
zero uncertainty. Asymptotic results condition on the selected smoothing
settings, without bandwidth or polynomial-degree selection uncertainty.
Plot bands apply within a panel, not jointly across all panels; asymptotic
simultaneous bands are not available. See `?np.plot` for details.

Use current dotted names such as `gradient.order`, `data.overlay` and
`data.rug`, and `B` for bootstrap replication counts where documented.
Unknown named arguments are rejected with a spelling hint where available.
The full package NEWS (`news(package = "npRmpi")`) lists the replacements
and family-specific changes. The [Gallery FAQ](https://jeffreyracine.github.io/gallery/faq.html)
provides further troubleshooting guidance.

## Joint polynomial-order and bandwidth selection

The same joint polynomial-order and bandwidth selection route is also available
here for local-polynomial-capable methods via
`search.engine = "nomad+powell"`, but this package-side vignette stays
operational rather than methodological.

If you want the recommended route without spelling out all of the LP tuning
arguments, use `nomad = TRUE`. This is a documented convenience preset, not a
generic optimizer alias: it fills only missing LP degree-search controls and
leaves compatible explicit overrides in place. This route uses the optional
NOMAD backend provided by the suggested package `crs`, so install `crs` first
if you want to use `nomad = TRUE` or
`search.engine = "nomad"`/`"nomad+powell"`.

```{r, purl = FALSE}
if (requireNamespace("crs", quietly = TRUE) &&
    utils::packageVersion("crs") >= package_version("0.15-41")) {
  npRmpi.init(nslaves = 1)

  set.seed(7)
  n <- 120
  x <- runif(n, -1, 1)
  y <- x + 0.4 * x^2 + rnorm(n, sd = 0.18)

  fit_nomad <- npreg(y ~ x, nomad = TRUE, degree.max = 1L, nmulti = 1L)
  fit_nomad$bws$nomad.shortcut

  bw_nomad_only <- npregbw(
    y ~ x,
    nomad = TRUE,
    search.engine = "nomad",
    degree.max = 1L,
    nmulti = 1L
  )
  bw_nomad_only$nomad.shortcut

  npRmpi.quit()
}
```

The same convenience entry point is available for the other LP-capable families:
`npcdens`, `npcdist`, `npplreg`, `npscoef`, and `npindex`, together with their
corresponding `*bw` constructors.

On current package sources, `spawn` is not the Windows entry path. There the
practical starting point is `attach` mode under `mpiexec`.

## Attach mode

Use attach mode when the MPI world is already launched via `mpiexec` and you
want `npRmpi.init(mode = "attach")` to be the startup entry point.

```{r, purl = FALSE}
library(npRmpi)
npRmpi.init(mode = "attach")

if (mpi.comm.rank(0L) == 0L) {
  set.seed(1)
  x <- runif(200)
  y <- sin(2 * pi * x) + rnorm(200, sd = 0.2)

  bw <- npregbw(y ~ x, regtype = "ll", bwmethod = "cv.ls")
  fit <- npreg(bws = bw)
  summary(fit)

  npRmpi.quit(mode = "attach")
}
```

For attach mode, keep `R_PROFILE_USER` and `R_PROFILE` cleared at launch time.
Profile startup belongs to a different mode and should not be mixed into
attach launches.

## Profile / manual-broadcast mode

Use profile mode when explicit startup and broadcast control are needed,
especially on larger or heterogeneous clusters. In this route,
`npRmpi.init()` is not the startup entry point; instead, ranks are launched
with `inst/Rprofile` (or `R_PROFILE_USER`) and then initialized explicitly
inside the script.

```{r, purl = FALSE}
mpi.bcast.cmd(np.mpi.initialize(), caller.execute = TRUE)

set.seed(1)
x <- runif(200)
y <- sin(2 * pi * x) + rnorm(200, sd = 0.2)
dat <- data.frame(y, x)

mpi.bcast.Robj2slave(dat)
mpi.bcast.cmd(
  bw <- npregbw(y ~ x, regtype = "ll", bwmethod = "cv.ls", data = dat),
  caller.execute = TRUE
)
mpi.bcast.cmd(
  fit <- npreg(bws = bw, data = dat),
  caller.execute = TRUE
)

summary(fit)
mpi.bcast.cmd(mpi.quit(), caller.execute = TRUE)
```

For profile mode, use exactly one startup profile source via `R_PROFILE_USER`
or a local `.Rprofile`, clear `R_PROFILE`, and avoid `--vanilla`.

## Practical guidance

- Start with `session` / `spawn` mode first on macOS and Linux.
- Start with `attach` mode first on Windows.
- Treat `attach` and `profile` as standard advanced workflows.
- Use `attach` when you want `npRmpi.init(mode = "attach")` to manage the
  worker loop for an already-launched MPI world.
- Use `profile` when you want startup and broadcasts to be explicit and
  script-controlled.
- Keep the attach and profile launch contracts distinct rather than mixing them.
- Keep the package-side vignette light; use longer website documentation for
  platform-specific install notes and cluster recipes.

## Where to go next

- `?npRmpi.init` and package README for current install notes
- demo scripts for `attach` and `profile` workflows
- <https://jeffreyracine.github.io/gallery/mpi_large_data.html> for the longer
  install/run guide
- <https://jeffreyracine.github.io/gallery/np_npRmpi.html> for the serial-to-MPI
  transition story
